\myTitle

### 2024-05-21 tex/body-text-of-paper.tex
add paragraph (overleaf)

A number of SNP markers from earlier discovery efforts \citep{clemento2011discovery} had already 
been validated as useful for GSI
and PBT in California \citep{clemento2014evaluation}. These markers were originally assayed
using TaqMan (Life Technologies Corp., Carlsbad, CA) probes or SNPtype assays (Standard BioTools Inc., South San Francisco, CA).
In order to type these markers using amplicon sequencing, we followed the methods in
\citet{campbell2015genotyping} and added the Illumina small RNA sequencing primer and 
read 2 sequencing primer to the SNPtype Specific Target Amplification Primer and Locus 
Specific Primers respectively.  We then dropped any loci that produced genotypes that were 
frequently inconsistent with genotypes produced with the SNPtype assays and any loci that 
did not amplify well or those that amplified so well that they were grossly over-represented
among the reads on the sequencer.  We 
used Freebayes \citep{garrison2012haplotype} to identify all variants on the fragment (not just the SNPs targeted by the original TaqMan assays).  This allows us to score each locus as a
microhaplotype \citep{baetscher2018microhaplotypes} when multiple variants occur in the sequence.  We filtered out variants
that were considered multinucleotide or complex variants by Freebayes. The resulting genetic markers are he

@@

We undertook a further discovery process to develop additional multiallelic
microhaplotype \citep{baetscher2018microhaplotypes} markers in California Chinook salmon.
\comment{NEED TO DESCRIBE. How did we look for these, etc. What constituted a
candidate to then follow up on with amplicon sequencing, etc.}

@@

The existing SNPtype assays and the additional microhaplotype markers were
developed prior to the advent of a well-assembled genome for Chinook salmon.
As a consequence, many of these markers had
been used previously without certainty about their location in the genome.
After designing and validating an amplicon-sequencing approach to type both the
existing SNPtype assays and the product of the additional microhaplotype
discovery process we identified the locations of these markers within the genome by mapping the
consensus sequences (roughly 300 bp) around the markers
to both the Chinook salmon version 1 (Otsh\_v1.0; \url{https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002872995.1/}) and version 2  (Otsh\_v2.0; \url{https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_018296145.1/}) genome assemblies,
using both the Burrows-Wheeler Aligner \citep{bwa-mem2009} and the BLAST-like Alignment tool \citep{kent2002blat}.
A similar mapping exercise was performed for the remaining markers so that we can
report their locations in both Otsh\_v1.0 and Otsh\_v2.0.

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon 
in California that includes multiallelic gene regions with high variability for 
relationship inference and gene regions identified in whole-genome sequence data 
for increased power for population identification. We show how this set of markers
provides sufficient power for highly accurate identification of parent and offspring
pairs and full siblings in California Chinook salmon, and also allows near-perfect 
identification of individuals to population or genetic group of origin. This set of 
genetic markers lays the foundation for a comprehensive genetic monitoring and evaluation effort
that facilitates multiple types of inference and is flexible and extensible.
When, and if, sequencing costs drop sufficiently, it may be possible to include
all different populations of a single species within a single standardized reference baseline
that performs equally well at broad and regional scales. A statistical approach for population
assignment from low coverage whole genome sequences appeared recently
\citep{desaixINPRESSpopulation}, and its authors noted that such a data type could
be ideal for producing reference baselines simultaneously applicable to broad scale
coverage and high resolution within sub-regions.   For the present, however,
baselines tailored to specific regions are essential for regional management questions.
It should be noted that recent

>>>>

A number of SNP markers from earlier discovery efforts \citep{clemento2011discovery} had already 
been validated as useful for GSI
and PBT in California \citep{clemento2014evaluation}. These markers were originally assayed
using TaqMan (Life Technologies Corp., Carlsbad, CA) probes or SNPtype assays (Standard BioTools Inc., South San Francisco, CA).
In order to type these markers using amplicon sequencing, we followed the methods in
\citet{campbell2015genotyping} and added the Illumina small RNA sequencing primer and 
read 2 sequencing primer to the SNPtype Specific Target Amplification Primer and Locus 
Specific Primers respectively.  We then dropped any loci that produced genotypes that were 
frequently inconsistent with genotypes produced with the SNPtype assays and any loci that 
did not amplify well or those that amplified so well that they were grossly over-represented
among the reads on the sequencer.  We 
used Freebayes \citep{garrison2012haplotype} to identify all variants on the fragment (not just the SNPs targeted by the original TaqMan assays).  This allows us to score each locus as a
microhaplotype \citep{baetscher2018microhaplotypes} when multiple variants occur in the sequence.  We filtered out variants
that were considered multinucleotide or complex variants by Freebayes. The resulting genetic markers are hereafter referred to as 'SNPlicons'.

@@

We undertook a further discovery process to develop additional multiallelic
microhaplotype \citep{baetscher2018microhaplotypes} markers in California Chinook salmon. A total of 125 microhaplotype markers were developed for Chinook salmon 
\comment{NEED TO DESCRIBE. How did we look for these, etc. What constituted a
candidate to then follow up on with amplicon sequencing, etc.}

@@

The SNPlicons targets and the additional microhaplotype markers were
developed prior to the advent of a well-assembled genome for Chinook salmon.
As a consequence, many of these markers had
been used previously without certainty about their location in the genome.
After designing and validating an amplicon-sequencing approach to type both the
existing SNPtype assays and the product of the additional microhaplotype
discovery process we identified the locations of these markers within the genome by mapping the
consensus sequences (roughly 300 bp) around the markers
to both the Chinook salmon version 1 (Otsh\_v1.0; \url{https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002872995.1/}) and version 2  (Otsh\_v2.0; \url{https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_018296145.1/}) genome assemblies,
using both the Burrows-Wheeler Aligner \citep{bwa-mem2009} and the BLAST-like Alignment tool \citep{kent2002blat}.
A similar mapping exercise was performed for the remaining markers so that we can
report their locations in both Otsh\_v1.0 and Otsh\_v2.0.

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon 
in California that includes multiallelic gene regions with high variability for 
relationship inference and gene regions identified in whole-genome sequence data 
for increased power for population identification. We show how this set of markers
provides sufficient power for highly accurate identification of parent and offspring
pairs and full siblings in California Chinook salmon, and also allows near-perfect 
identification of individuals to population or genetic group of origin. This set of 
genetic markers lays the foundation for a comprehensive genetic monitoring and evaluation effort
that facilitates multiple types of inference and is flexible and extensible.
When, and if, sequencing costs drop sufficiently, it may be possible to include
all different populations of a single species within a single standardized reference baseline
that performs equally well at broad and regional scales. A statistical approach for population
assignment from low coverage whole genome sequences appeared recently
\citep{desaixINPRESSpopulation}, and its authors noted that such a data type could
be ideal for producing reference baselines simultaneously applicable to broad scale
coverage and high resolution within sub-regions.   For the present, however,
baselines tailored to specific regions are essential for regional management questions.

Accurate identification of the distinct ecotypes of Chinook salmon in the California Central Valley is of utmost importance for designing and implementing effective conservation and management actions. This has been challenging, given the recent common ancestors of these ecotypes and ongoing migration between subbasins. We describe the first set of genetic markers that identifies the late-fall-run salmon ecotype, which occurs only in the CCV and shares a genomic background with the more common CCV fall-run salmon ecotype. Moreover, we show how the FRH spring-run lineage is easily identifiable through a combination of traditional GSI and the characterization of functional genetic markers in the RoSA. Finally, although previous work has described GSI capabilities that distinguish the natural-spawning CCV spring-run lineages from each other and their fall-run counterparts with moderately high accuracy, we demonstrate near complete accuracy in distinguishing these 'stocks'. Moreover, the few fish that are apparently misidentified (Fig. 3) are likely to be primarily migrants and not true misidentifications. For example, the three fish that were field-characterized as spring-run from Butte Creek, but are genetically identified as winter-run salmon are clearly carry the winter-run genomic background and could not realistically be misidentified on the basis of the genetic data. This marker set and baseline reference dataset also provides unprecedented power for identifying fish from not only the other ESUs in California, but even the individual populations within those ESUs (Fig. 3).
Additionally, previous genetic marker sets for California Chinook salmon had sufficient power for inferring parent-offspring relationships, but only when both parents were sampled and genotyped. This marker set considerably increases the capacity for relationship inference in California salmon, by providing sufficient power for parent-offspring relationship inference, when only one parent was sampled, as well as identifying pairs of full siblings when no parents are sampled. Groups of full siblings larger than two are identified even more accurately than expressed in Figure 4.
